The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2013-06-23 script/de/chapters/numbertheory.tex
CT Script: Minor changes.

\begin{definition}\label{def-zth-remainder} \index{Restklasse}
Als {\bf Restklasse $r$ modulo $m$} bezeichnet man alle ganzen Zahlen $a$, die
bei der Division durch $m$ denselben Rest $r$ haben.
\end{definition}
\newpage
\begin{example}{:}
%\begin{itemize}
\begin{compactitem}
\item[] Restklasse $0$ modulo $4 = \{ x | x = 4*n; \; n \in \mathbb{Z} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] Restklasse $3$ modulo $4 = \{ x | x = 4*n + 3;\; n \in \mathbb{Z} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
%\end{itemize}
\end{compactitem}
\end{example}
Da modulo $m$ nur die Reste $0, 1, 2, \cdots, m-1$ möglich sind, rechnet die modulare Arithmetik in endlichen Mengen. 
Zu jedem Modul $m$ gibt es genau $m$ Restklassen.

@@

\begin{satz}\label{thm-zth-div}
$a \equiv b$ (mod $m$) gilt genau dann,  wenn die Differenz $(a - b)$ durch $m$
teilbar ist, also wenn ein $q\in \mathbf{Z}$ existiert mit $ (a-b)=q*m.$\footnote{%
Die obige Äquivalenz gilt nur für die Differenz $(a - b)$, nicht für die
Summe $(a + b)$!

@@

Daraus ergibt sich: Wenn $m$ die Differenz teilt, gibt es eine ganze Zahl $q$, so
dass gilt: $a = b + q*m$.
Alternativ zur Kongruenzschreibweise kann man auch die Teilbarkeitsschreibweise
verwenden: $m | (a - b)$.

@@

%\newpage
\begin{example}{:} \\
Ist $69.993$ durch $7$ teilbar?\\
Da die Zahl in eine Differenz zerlegt werden kann, bei der einfach zu ersehen
ist, dass jeder Operand durch $7$ teilbar ist, ist auch die Differenz durch $7$
teilbar: $69.993 = 70.000 - 7$.
\end{example}

@@

Diese \glqq modulare'', also auf die Divisionsreste bezogene Arithmetik ist die
Basis der asymmetrischen Verschlüsselungsverfahren.
Kryptographische Berechnungen spielen sich also nicht wie das Schulrechnen
unter den reellen Zahlen ab, sondern unter Zeichenketten begrenzter Länge,
das heißt unter positiven ganzen Zahlen, die einen gewissen Wert nicht
überschreiten dürfen.
Aus diesem und anderen Gründen wählt man sich eine große Zahl $m$ und \glqq rechnet
modulo $m$'', das heißt, man ignoriert ganzzahlige Vielfache von $m$ und rechnet
statt mit einer Zahl nur mit dem Rest bei Division dieser Zahl durch $m$.
Dadurch bleiben alle Ergebnisse im Bereich von $0$ bis $m-1$.

>>>>

\begin{definition}\label{def-zth-remainder} \index{Restklasse}
Als {\bf Restklasse $r$ modulo $m$} bezeichnet man alle ganzen Zahlen $a$, die
bei der Division durch $m$ denselben Rest $r$ haben.
\end{definition}
\newpage
\begin{example}{ von Restklassen:}
%\begin{itemize}
\begin{compactitem}
\item[] Restklasse $0$ modulo $4 = \{ x | x = 4*n; \; n \in \mathbb{Z} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] Restklasse $3$ modulo $4 = \{ x | x = 4*n + 3;\; n \in \mathbb{Z} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
%\end{itemize}
\end{compactitem}
\end{example}
Da modulo $m$ nur die Reste $0, 1, 2, \cdots, m-1$ möglich sind, rechnet die modulare Arithmetik in endlichen Mengen. 
Zu jedem Modul $m$ gibt es genau $m$ Restklassen.

Das Ergebnis der Modulo-Operation lässt sich so ausdrücken:
$~a \bmod{m} ~ =  ~ a - m * \lfloor a/m \rfloor$\\

@@

\vskip +20 pt
\begin{satz}\label{thm-zth-div}
$a \equiv b$ (mod $m$) gilt genau dann,  wenn die Differenz $(a - b)$ durch $m$
teilbar ist, also wenn ein $q\in \mathbf{Z}$ existiert mit $ (a-b)=q*m.$\footnote{%
Die obige Äquivalenz gilt nur für die Differenz $(a - b)$, nicht für die
Summe $(a + b)$!

@@

Daraus ergibt sich: Wenn $m$ die Differenz teilt, gibt es eine ganze Zahl $q$, so
dass gilt: $a = b + q*m$.
Alternativ zur Kongruenzschreibweise kann man auch die Teilbarkeitsschreibweise
verwenden: $m | (a - b)$.\\

@@

%\newpage
\begin{example}{:} \\
Ist $69.993$ durch $7$ teilbar?\\
Da die Zahl in eine Differenz zerlegt werden kann, bei der einfach zu ersehen
ist, dass jeder Operand durch $7$ teilbar ist, ist auch die Differenz durch $7$
teilbar: $69.993 = 70.000 - 7$.\\
\end{example}

@@

Die \glqq modulare'', also auf die Divisionsreste bezogene Arithmetik ist die
Basis der asymmetrischen Verschlüsselungsverfahren.
Kryptographische Berechnungen spielen sich also nicht wie das Schulrechnen
unter den reellen Zahlen ab, sondern unter Zeichenketten begrenzter Länge,
das heißt unter positiven ganzen Zahlen, die einen gewissen Wert nicht
überschreiten dürfen.
Aus diesem und anderen Gründen wählt man sich eine große Zahl $m$ und \glqq rechnet
modulo $m$'', das heißt, man ignoriert ganzzahlige Vielfache von $m$ und rechnet
statt mit einer Zahl nur mit dem Rest bei Division dieser Zahl durch $m$.
Dadurch bleiben alle Ergebnisse im Bereich von $0$ bis $m-1$.

### 2013-06-23 script/en/chapters/numbertheory.tex
CT Script: Minor changes.

\newpage
\begin{example}{:}
\begin{compactitem}
\item[] Remainder class $0$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n; \; n \in \mathbb{Z} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] Remainder class $3$ modulo 
        4 = \\ 
        \strut\quad $\{ x | x = 4 * n + 3;\; n \in \mathbb{Z} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
\end{compactitem}
\end{example}
As only the remainders $0, 1, 2, \cdots, m-1$ are possible for division modulo $m$, modular arithmetic works with finite sets. 
For each modulo $m$ there are precisely $m$ remainder classes.

@@

\begin{theorem}\label{thm-zth-div}
$a \equiv b$ (mod $m$) if and only if, the difference $(a - b)$ is divisible by
$m$, i.e. if $q\in \mathbb{Z}$ exists with $ (a-b)=q*m.$\footnote{%
The above equivalence does apply only to the difference $(a - b)$, not to the
sum $(a + b)$!

@@

Therefore: If $m$ divides the difference, there exists a whole number
$q$ such that: $a = b + q*m$.
As an alternative to the congruence notation, we can also use the divisibility
notation: $m | (a - b)$.

@@

\begin{example}{:}
Is $69,993$ divisible by $7$? \\
The number can be written in the form of a difference in which it is clear that
each operand is divisible by $7$: 
$69,993 = 70,000 - 7$. Therefore, the difference is also divisible by $7$.
\end{example}

@@

This ``modular'' arithmetic (based on division remainders) forms the basis of asymmetric encryption procedures. 
Cryptographic calculations are therefore not based on real numbers, as the calculations you performed at school, 
but rather on character strings with a limited length, in other words on positive whole numbers that cannot exceed 
a certain value. This is one of the reasons why we choose a large number $m$ and ``calculate modulo $m$''. 
That is, we ignore whole-number multiples of $m$ and, rather than working with a number, we only work with 
the remainder when this number is divided by $m$. The result is that all results are in the range $0$ to $m-1$.

@@

Within the remainder tree you get -- at the end -- $(P \bmod ({m_{i}^{2}}) ) / m_{i}$ for all $i$.\footnote{%
It would not make sense to calculate $(P \bmod m_{i} ) / m_{i}$, as 
$m_{i} | P$, so $P/m_{i}$ is always a whole number, which means  
$(P \bmod m_{i} )$ is always $= 0$.

Sample with very small moduli:\\
$ m_{1} = 2*3 = 6;~~~ m_{2} = 2*7 = 14;~~~ P=6*14=84 $\\
$ P \bmod m_{1} = 84 \bmod 6 = 0;~~~ P \bmod {m_{1}^{2}} = 84 \bmod 36 = 12 $\\
$ P \bmod m_{2} = 84 \bmod 14 = 0;~~~ P \bmod {m_{2}^{2}} = 84 \bmod 196 = 84 $\\
$gcd_{1} = gcd(12/6, 6) = gcd(2, 6) = 2 $\\
$gcd_{2} = gcd(84/14, 14) = gcd(6, 14) = 2 $

@@

$$(P \bmod{m_{i}^{2}})/ m_{i} - P / m_{i} \equiv 0   \bmod{m_{i}}~~~ \Leftrightarrow ~~~  m_{i} ~~ | ~~ ((P \bmod{m_{i}^{2}})/ m_{i} - P / m_{i})$$~~~\footnote{%
According to the definition of modulus (see theorem \ref{thm-zth-div}) and
division it is:
$a \bmod{b} ~~ \Longleftrightarrow ~~ a - b * \lfloor a/b \rfloor$\\
So $P \bmod{m_{i}^{2}}$ can be written as $ P-m_{i}^{2} \lfloor P/m_{i}^{2} \rfloor$.
}

>>>>

\newpage
\begin{example}{ of remainder classes RC:}
\begin{compactitem}
% \strut\quad
\item[] RC $0$ mod $ 4 = \{ x | x = 4 * n; \; n \in \mathbb{Z} \} = \{ \dots, -16, -12, -8, -4, 0, 4, 8, 12, 16, \dots \}$
\item[] RC $3$ mod $ 4 = \{ x | x = 4 * n + 3;\; n \in \mathbb{Z} \} = \{ \dots, -13, -9, -5, -1, 3, 7, 11, 15, \dots \}$
\end{compactitem}
\end{example}
As only the remainders $0, 1, 2, \cdots, m-1$ are possible for division modulo $m$, modular arithmetic works with finite sets. 
For each modulo $m$ there are precisely $m$ remainder classes.

The result of the modulo operation can be formulated as:
$~a \bmod{m} ~ =  ~ a - m * \lfloor a/m \rfloor$\\

@@

\vskip +20 pt
\begin{theorem}\label{thm-zth-div}
$a \equiv b$ (mod $m$) if and only if, the difference $(a - b)$ is divisible by
$m$, i.e. if $q\in \mathbb{Z}$ exists with $ (a-b)=q*m.$\footnote{%
The above equivalence does apply only to the difference $(a - b)$, not to the
sum $(a + b)$!

@@

Therefore: If $m$ divides the difference, there exists a whole number
$q$ such that: $a = b + q*m$.
As an alternative to the congruence notation, we can also use the divisibility
notation: $m | (a - b)$.\\

@@

\begin{example}{:}
Is $69,993$ divisible by $7$? \\
The number can be written in the form of a difference in which it is clear that
each operand is divisible by $7$: 
$69,993 = 70,000 - 7$. Therefore, the difference is also divisible by $7$.\\
\end{example}

@@

The ``modular'' arithmetic (based on division remainders) forms the basis of
asymmetric encryption procedures. 
Cryptographic calculations are therefore not based on real numbers, as the
calculations you performed mostly at school, but rather on character strings
with a limited length, in other words on positive whole numbers that cannot
exceed a certain value. This is one of the reasons why we choose a large number
$m$ and ``calculate modulo $m$''. 
That is, we ignore whole-number multiples of $m$ and, rather than working with
a number, we only work with the remainder when this number is divided by $m$.
The result is that all results are in the range $0$ to $m-1$.

@@

Within the remainder tree you get -- at the end -- $(P \bmod ({m_{i}^{2}}) ) / m_{i}$ for all $i$.\footnote{%
It would not make sense to calculate modulo $m_{i}$ instead of $m_{i}^{2}$ on the
left gcd, which would mean to use $(P \bmod m_{i} ) / m_{i}$, as 
$m_{i} | P$, so $P/m_{i}$ is always a whole number, which means  
$(P \bmod m_{i} )$ is always $= 0$.

\noindent Sample with very small moduli:\\
$ m_{1} = 2*3 = 6;~~~ m_{2} = 2*7 = 14;~~~ P=6*14=84 $\\
$ P \bmod m_{1} = 84 \bmod 6 = 0;~~~ P \bmod {m_{1}^{2}} = 84 \bmod 36 = 12 $\\
$ P \bmod m_{2} = 84 \bmod 14 = 0;~~~ P \bmod {m_{2}^{2}} = 84 \bmod 196 = 84 $\\
$gcd_{1} = gcd(12/6, 6) = gcd(2, 6) = 2 $\\
$gcd_{2} = gcd(84/14, 14) = gcd(6, 14) = 2 $

@@

$$(P \bmod{m_{i}^{2}})/ m_{i} - P / m_{i} \equiv 0   \bmod{m_{i}}~~~ \Leftrightarrow ~~~  m_{i} ~~ | ~~ ((P \bmod{m_{i}^{2}})/ m_{i} - P / m_{i})$$~~~\footnote{%
Using the modulus operation (definition \ref{def-zth-remainder} at page
\pageref{def-zth-remainder}) and division it is:
$~a \bmod{b} ~ =  ~ a - b * \lfloor a/b \rfloor$\\
So $P \bmod{m_{i}^{2}}$ can be written as $ P-m_{i}^{2} \lfloor P/m_{i}^{2} \rfloor$.
}
